---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  size = "footnotesize"
)
```

# AdaptiveDesignSim

<!-- badges: start -->
<!-- [![R-CMD-check](https://github.com/dcr-unibe-ch/AdaptiveDesignSim/workflows/R-CMD-check/badge.svg)](https://github.com/dcr-unibe-ch/AdaptiveDesignSim/actions) -->
<!-- badges: end -->

An R package for simulations of clinical trials with adaptive designs.


## Installation

The package can be installed from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("dcr-unibe-ch/AdaptiveDesignSim")
```

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
devtools::load_all()
options(width = 200)
```

## Simulation

A single trial with one or more interim analyses can be simulated using `ADsim`.
The required options are the sample size in each arm (*n01*),
	the number and potentially time point of interim analyses (*nia*, *tia*).
	the distribution (*simfun*) and paremeters in each arm (*par01*), 
	and the estimation function (*estfun*).

Let's assume

- a trial with a target final sample size of 200 (1:1 allocation),
- a 'good' binary outcome, which is expected to occur in 20% of the control patient,
- a relevant effect would be an absolute increase by 20%,
- a risk difference as effect measure using a Wald-type CI and z-test.

We would like to do one interim analysis at half of the patients.
Single simulations under the null and alternative can the be done like that:

```{r}
set.seed(12)

simH0<-ADsim(n01 = c(100, 100), nia = 1, 
	par01 = c(p0=0.2, p1=0.2), simfun = "binom",
	estfun = "rd_wald", 
	direct = "higher")

simH1<-ADsim(n01 = c(100, 100), nia = 1, 
	par01 = c(p0=0.2, p1=0.4), simfun = "binom",
	estfun = "rd_wald", 
	direct = "higher")
	
cbind(simH0,simH1)
```

`ADsim` returns a numeric vector with the true point estimate
(*true_pe*, a risk difference in this example), 
and---for each interim stage (prefix *iax_*) and the final stage (prefix *fa_*)---the

- simulated number of successes for both arms (suffix *x0*, *x1*),
- sample size for both arms (suffix *n0*, *n1*),
- simulated probabilities for both arms (suffix *p0*, *p1*),
- simulated point estimate (suffix *pe*, a risk difference in this example),
- two-sided *cilevel*% confidence intervals (suffix *lci* and *uci*), and 
- z-statistic (suffix *z*).

Simulations can be repeated and stacked to generate a data frame that can then be used with `ADeval`.

```{r}
set.seed(12)

simdatH0 <- lapply(1:1000, function(i) 
	ADsim(n01 = c(100, 100), nia = 1, 
	par01 = c(p0=0.2, p1=0.2), simfun = "binom",
	estfun = "rd_wald", 
	direct = "higher"))
simdatH0 <- do.call(rbind, simdatH0)

simdatH1 <- lapply(1:1000, function(i) 
	ADsim(n01 = c(100, 100), nia = 1, 
	par01 = c(p0=0.2, p1=0.4), simfun = "binom",
	estfun = "rd_wald", 
	direct = "higher"))
simdatH1 <- do.call(rbind, simdatH1)

```

### User-defined estimation function

Estimation functions can be specified via **estfun**.
	There are five defaults, standard Wald-type approximation 
	for the risk difference (*wald_rd*) and risk ratio (*wald_rr*),
	the score-based method for the risk difference suggested by Newcombe (*score_rd*), 
	the score-based method for the risk ratio according to Koopman (*score_rr*), 
	and Student's t-test for the mean difference (*md_t*).

Alternatively, any function can be defined, which takes hte output from **simfun** and *cilevel* as arguements.

The output of the function must be a (named) vector with point estimate, lower confidence limit,
	upper confidence limit and z-statistic.
	It can also contain further elements (e.g. estimates for each group), which are added to the output of `ADsim`.
	
If a user-specified function is given in **estfun**, option **truefun** has also to be used. 
	It must be a function with argument *par01* that specifies how the true effect is calculated,
		e.g. as *diff(par01)* for the risk difference.

This example would reproduce the *wald_rd* option.

```{r, eval =FALSE}

efun<-function(out,cilevel) {
	xs<-unlist(lapply(out,sum))
	ns<-unlist(lapply(out,length))
	pe<-diff(xs/ns)
	se<-sqrt(sum(xs/ns*(1-xs/ns)/ns))
	lci<-pe - qnorm(1-(1-cilevel)/2)*se
	uci<-pe + qnorm(1-(1-cilevel)/2)*se
	z<-(pe)/se
	res<-c(pe,lci,uci,z)
	names(res)<-c("pe","lci","uci","z")
	return(res)
}

tfun <- function(x) diff(x)

ADsim(n01 = c(100,100), nia = 1,
      par01 = c(0.2,0.4), simfun = "binom",
      estfun = efun, truefun = tfun,
      direct = "higher")

```
	
### Different outcome types

Continuous outcomes can be specified like that: 

```{r, eval=FALSE}

par01<-data.frame(matrix(c(5,1,5.2,1),2,2))
rownames(par01)<-c("mean","sd")

ADsim(n01 = c(100,100), nia = 1, 
      par01 = par01, simfun="norm",
      estfun = "md_t",
      direct = "higher")

```

Note that the *par01* needs rows named *mean* and *sd*.

Other outcome types can also be included using specific functions for **simfun**, **estfun** and **truefun**.

**simfun** has to take *par01* as argument, and produce a list of length two with the simulated outcomes for control and experimental group.

For example, a Poisson distribution with expected counts of 4 and 3 in control and experimental group and low counts being better:

```{r, eval=FALSE}

sfun<-function(n01,par01) {
	out0<-rpois(n = n01[1], lambda = par01[1])
	out1<-rpois(n = n01[2], lambda = par01[2])
	return(list(out0,out1))
}

efun<-function(out,cilevel) {
	
	ns<-unlist(lapply(out,length))
	xs<-unlist(lapply(out,mean))
	
	m1<-glm(c(out[[1]],out[[2]]) ~ c(rep(0,ns[1]),rep(1,ns[2])),
	        poisson(link = "log"))
	
	z<-summary(m1)$coef[2,"z value"]
	
	res<-c(xs,m1$coef[2],confint.default(m1, level = cilevel)[2,],z)

	names(res)<-c("m0","m1","pe","lci","uci","z")
	return(res)
}

tfun<-function(p01) diff(log(par01))

ADsim(n01 = c(100,100), nia = 1,
	par01 = c(4,3), simfun=sfun,
	estfun = efun, truefun = tfun,
	direct="lower")

```


## Evaluation

Repeated simulations can be evaluated using function `ADeval`. 

Thresholds for stopping at each interim analysis have to be defined via **zcutoff**,
 a matrix with one row per interim analysis and a column for stopping for futility and efficacy.
	A stop for futility will be assumed
	if the observed Z-statistic at interim is lower than the threshold in column 1, 
	a stop for efficacy if the observed Z-statistic at interim is higher than the threshold in column 2.
	*NA* would mean no stopping at the respective interim analysis.
	
Let's assume we have one interim analysis (as specified in the `ADsim` above) and would like to stop 
	for futility if Z<0 and for efficacy if Z>2:

```{r}
zcutoff<-matrix(c(0,2),1,2)
zcutoff
```

```{r}
resH0<-ADeval(simdat = simdatH0, zcutoff = zcutoff)

resH1<-ADeval(simdat = simdatH1, zcutoff = zcutoff)

head(resH0$rawdata)

cbind(resH0$opchar,resH1$opchar)

```

`ADeval` returns a list with the evaluations for each repetition (*rawdata*) and 
	the summarized operating characteristics (*opchar*).

*rawdata* is a data frame with the elements from `ADsim` plus 

- prefix *obs_*: the observed sample size (*n0*, *n1*), effect (*pe*) and confidence interval (*lci*, *uci*) at study end 
	(which could be at an interim or the final stage),
	and potentially further variables included in `ADsim::estfun`.
- *tstop*: the time point of the first stop for either futility or efficacy,	
	and of the first stop for futility and efficacy (*tstop_fut* and *tstop_eff*).
- *anystop*: an indicator whether there was any stop, and a stop for futility or efficacy (*anystop_fut* and *anystop_eff*).

*opchar* is a numeric vector with operating characteristics over all simulated trials including the

- prefix *av_*: average estimates in each group, effect (point estimate, *pe*) and 95% confidence limits (*lci*, *uci*).
- *psig*: probability for a significant trial (the power or type I error, depending on the assumptions).
- *pstop*: probability for any stop and a stop for futility or efficacy (*pstop_fut*, *pstop_eff*), 
	overall and at each interim analysis (suffix *iax*).
- average, minimum and maximum total sample size (av_n, min_n, max_n).
- *bias*: average of estimate minus true value).
- *sd*: standard deviation of the point estimate.
- *mse*: mean squared error, average of the squared difference between estimate and true value).
- *coverage*: proportion of confidence intervals including the true value.

## shinyApp

A shinyApp can be launched locally in RStudio via `launch_ADSim_app()`.



