---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  size = "footnotesize"
)
```

# AdaptiveDesignSim

<!-- badges: start -->
[![R-CMD-check](https://github.com/dcr-unibe-ch/AdaptiveDesignSim/workflows/R-CMD-check/badge.svg)](https://github.com/dcr-unibe-ch/AdaptiveDesignSim/actions)
<!-- badges: end -->

An R package for simulations of clinical trials with adaptive designs,


## Installation

The package can be installed from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("dcr-unibe-ch/AdaptiveDesignSim")
```

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
devtools::load_all()
options(width = 200)
```

## Binary outcome

A single trial with one or more interim analyses can be simulated using `ADsimfun`.
The required options are the sample size in each arm (*n01*),
  the proportion of the outcome in each arm (*p01*), 
  and the number of interim analyses (*nia*).

```{r}
ads<-ADsimfun(n01 = c(100,100), p01 = c(0.2, p1=0.4), nia = 1, alpha=0.025)
data.frame(ads)
```


`ADsimfun` returns a numeric vector with the
  true proportions in both arms and point estimate (*true_p0*, *true_p1*, *true_pe*), and,
  for each interim stage (prefix *iax_*) and the final stage (prefix *fa_*),
  the simulated number of successes, sample size and 
  observed probabilities for both arms (suffix *_x0*, *_x1*, *_n0*, *_n1*, *_p0*, *_p1*),
  point estimate (suffix *_pe*), confidence interval (suffix *_lci* and *_uci*) and z-statistic (*_z*).

Further options for `ADsimfun` are

- the time point of the interim analysis (*pia*, as number of participants, equally spaced by default),
- the significance level for the final test (*alpha*, 0.025 by default),
- the effect measure (*effm*, risk difference or ratio),
- the direction of the effect (*direct* with **lower** is better or **higher** is better), and 
- the type of confidence interval used for the analysis (*cimethod*, **Wald** or **score**).

Simulations can be repeated and stacked to generate a data frame that can be used with `ADevalfun`.

```{r}
simdat <- lapply(1:1000, function(i) 
	ADsimfun(n01 = c(100,100), p01 = c(0.2, p1=0.4), nia = 1, alpha = 0.025, effm="rd", direct="lower"))
simdat <- do.call(rbind, simdat)
```

For the `ADevalfun` thresholds for stopping at each interim analysis have to be defined via *zcutoff*,
 a data frame with one row per interim analysis and a column for stopping for futility and efficacy.
	A stop for futility will be assumed
	if the observed Z-statistic at interim is lower than the threshold in column 1, 
	a stop for efficacy if the observed Z-statistic at interim is higher than the threshold in column 2.
	NA would mean no stopping at the respective interim analysis.
	
Let's assume we have one interim analysis (as specified in the `ADsimfun` above) and would like to stop 
	for futility if Z<0 and for efficacy if Z>2:

```{r}
zcutoff<-t(data.frame(c(0,2)))
zcutoff
```

```{r}
sr<-ADevalfun(simdat = simdat, zcutoff = zcutoff)
data.frame(sr$opchar)

```

`ADevalfun` returns a list with the evaluations for each repetition (*rawdata*) and 
	the summarised operating characteristics (*opchar*).

*rawdata* is a data frame with the elements from `ADsimfun` plus 

- the final proportions point estimate and confidence interval at study end (which could be the interim or final stage),
- the time point of the first stop (*tstop*, could be for futility or efficacy),	
	and of the first stop for futility and efficacy (*tstop_fut* and *tstop_eff*),
- the final sample size (*neff*), and
- an indicator whether there was any stop (*anystop*) and a stop for futility or efficacy (*anystop_fut* and *anystop_eff*).

*opchar* is a numeric vector with operating characteristics over all simulated trials including the
- average estimates in each group, effect (point estimate, *pe*) and (1-alpha) confidence limits (*lci*, *uci*).
- probability for a significant trial (*psig*, the power or type I error, depending on the assumptions).
- probability for any stop (*pstop*) and a stop for futility or efficacy (*pstop_fut*, *pstop_eff*).
- average, minimum and maximum sample size (avn, minn, maxn).
- bias (average of estimate minus true value).
- standard deviation (*sd*) of the point estimate.
- mean squared error (*mse*, average of the squared difference between estimate and true value), and 
- coverage (proportion of confidence intervals including the true value).

