---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  size = "footnotesize"
)
```

# AdaptiveDesignSim

<!-- badges: start -->
<!-- [![R-CMD-check](https://github.com/dcr-unibe-ch/AdaptiveDesignSim/workflows/R-CMD-check/badge.svg)](https://github.com/dcr-unibe-ch/AdaptiveDesignSim/actions) -->
<!-- badges: end -->

An R package for simulations of clinical trials with adaptive designs.


## Installation

The package can be installed from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("dcr-unibe-ch/AdaptiveDesignSim")
```

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
devtools::load_all()
options(width = 200)
```

## Simulation

A single trial with one or more interim analyses can be simulated using `ADsim`.
The required options are the sample size in each arm (**n01**),
  the proportion of the outcome in each arm (**p01**), 
  and the number of interim analyses (**nia**).
  
Further optional arguments for `ADsim` are

- the time point of the interim analysis (**tia**, as fraction of the total, equally spaced by default),
- the estimation function (**estfun**, either a user-specified function 
	or a string with one of the four defaults (wald_rd, wald_rr, score_rd, score_rr),
- the two-sided confidence level for the confidence interval 
	if one of the defaults was used for *estfun*,
- a function to calculate the true point estimate if a user-specified *estfun* was used (**truefun**),	
- the direction of the effect (**direct** with *lower* is better or *higher* is better).

Let's assume

- a trial with a target final sample size of 200 (1:1 allocation),
- a 'bad' binary outcome, which is expected to occur in 40% of the control patient,
- a relevant effect would be a decrease by 20%,
- a risk difference as effect measure using Wald-type CI and test with a one-sided alpha of 0.025.

We would like to do one interim analysis at half of patients
Single simulation under the null and alternative can the be done like that:

```{r}
set.seed(12)

simH0<-ADsim(n01 = c(100, 100), p01 = c(0.4, p1=0.4), nia = 1, direct = "lower")

simH1<-ADsim(n01 = c(100, 100), p01 = c(0.4, p1=0.2), nia = 1, direct = "lower")

cbind(simH0,simH1)
```


`ADsim` returns a numeric vector with

- the true proportions in both arms (*true_p0*, *true_p1*).
- the true point estimate (*true_pe*), which is either a risk difference or a log risk ratio (depending on option **estfun**).
  
And for each interim stage (prefix *iax_*) and the final stage (prefix *fa_*) the

- simulated number of successes for both arms (suffix *x0*, *x1*),
- sample size for both arms (suffix *n0*, *n1*),
- simulated probabilities for both arms (suffix *p0*, *p1*),
- simulated point estimate (suffix *pe*, either the risk difference or the log risk ratio),
- two-sided *cilevel*% confidence intervals (suffix *lci* and *uci*), and 
- z-statistic (suffix *z*).

Simulations can be repeated and stacked to generate a data frame that can then be used with `ADeval`.

```{r}
set.seed(12)

simdatH0 <- lapply(1:1000, function(i) 
	ADsim(n01 = c(100,100), p01 = c(0.2, p1=0.2), nia = 1, 
	      direct="lower"))
simdatH0 <- do.call(rbind, simdatH0)

simdatH1 <- lapply(1:1000, function(i) 
	ADsim(n01 = c(100,100), p01 = c(0.4, p1=0.2), nia = 1,
	      direct="lower"))
simdatH1 <- do.call(rbind, simdatH1)

```

Optionally a different effect estimate or estimation function can be specified via **estfun**.
	There are four defaults, standard Wald-type approximation 
	for the risk difference (*wald_rd*) and risk ratio (*wald_rr*),
	the score-based method for the risk difference suggested by Newcombe (*score_rd*) and 
	the score-based method for the risk ratio according to Koopman (*score_rr*).

Alternatively, any function can be defined with arguments

- *xs*: the number of successes in control and experimental group)
- *ns*: the number of patients in control and experimental group
- *cilevel*: the confidence level 

and a named vector as output with point estimate, lower confidence limit,
	upper confidence limit and z-statistic.

In that case, function **truefun** with argument *p01* has to be used to specify 
how the true effect should be calculated, e.g. as *diff(p01)* for the risk difference.

This example would reproduce the **wald_rd** option: 

```{r, eval=FALSE}

efun<-function(xs,ns,cilevel) {
	pe<-diff(xs/ns)
	se<-sqrt(sum(xs/ns*(1-xs/ns)/ns))
	lci<-pe - qnorm(1-(1-cilevel)/2)*se
	uci<-pe + qnorm(1-(1-cilevel)/2)*se
	z<-(-pe)/se
	res<-c(pe,lci,uci,z)
	names(res)<-c("pe","lci","uci","z")
	return(res)
}

ADsim(n01 = c(100,100), p01 = c(0.4,0.2), nia = 1, estfun = efun, truefun = function(x) diff(x))

```
	


## Evaluation

For the `ADeval` thresholds for stopping at each interim analysis have to be defined via **zcutoff**,
 a matrix with one row per interim analysis and a column for stopping for futility and efficacy.
	A stop for futility will be assumed
	if the observed Z-statistic at interim is lower than the threshold in column 1, 
	a stop for efficacy if the observed Z-statistic at interim is higher than the threshold in column 2.
	*NA* would mean no stopping at the respective interim analysis.
	
Let's assume we have one interim analysis (as specified in the `ADsim` above) and would like to stop 
	for futility if Z<0 and for efficacy if Z>2:

```{r}
zcutoff<-matrix(c(0,2),1,2)
zcutoff
```

```{r}
resH0<-ADeval(simdat = simdatH0, zcutoff = zcutoff)

resH1<-ADeval(simdat = simdatH1, zcutoff = zcutoff)

cbind(resH0$opchar,resH1$opchar)

```

`ADeval` returns a list with the evaluations for each repetition (*rawdata*) and 
	the summarized operating characteristics (*opchar*).

*rawdata* is a data frame with the elements from `ADsim` plus 

- the observed proportions, effect and confidence interval at study end, which could be at an interim or the final stage  (*obs_p0*, *obs_p1*, *obs_pe*, *obs_lci*, *obs_uci*),
- the time point of the first stop for either futility or efficacy (*tstop*),	
	and of the first stop for futility and efficacy (*tstop_fut* and *tstop_eff*),
- the final sample size (*neff*), and
- an indicator whether there was any stop (*anystop*) and a stop for futility or efficacy (*anystop_fut* and *anystop_eff*).

*opchar* is a numeric vector with operating characteristics over all simulated trials including the

- average estimates in each group, effect (point estimate, *pe*) and 95% confidence limits (*lci*, *uci*),
- probability for a significant trial (*psig*, the power or type I error, depending on the assumptions),
- probability for any stop (*pstop*) and a stop for futility or efficacy (*pstop_fut*, *pstop_eff*), overall and at each interim analysis (suffix *iax*),
- average, minimum and maximum sample size (avn, minn, maxn),
- bias (average of estimate minus true value),
- standard deviation (*sd*) of the point estimate,
- mean squared error (*mse*, average of the squared difference between estimate and true value), and 
- coverage (proportion of confidence intervals including the true value).

## shinyApp

A shinyApp can be launched locally in RStudio via `launch_ADSim_app()`.



